---
title: "fracdet quickstart"
author: "Constantino Antonio García Martínez"
date: "`r Sys.Date()`"
output: html_document
vignette: >
  %\VignetteIndexEntry{fracdet_quickstart}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r options,echo=FALSE}
# colorblind friendly palette
palette(c("#000000", "#E69F00", "#56B4E9", "#009E73",
          "#F0E442", "#0072B2", "#D55E00", "#CC79A7"))
library('knitr')
knitr::opts_chunk$set(
  fig.width = 7, fig.height = 4.7,
  fig.align = "center", fig.show = 'hold'
  )
```

## Introduction
__fracdet__ is an R package implementing a simple, yet realistic,
fractal-deterministic model that may be able to capture key features of 
complex systems with long  term correlations. We assume that the observed 
signal $Y$ is the result of the superposition of a stochastic series $B$ and 
some band-limited deterministic signal $x$:

$$Y[n] = x[n] + B[n].$$
 
__fracdet__ also provides a method that, under the assumptions of this model,
permits the characterization of the fractal component and the  estimation of the
deterministic components of the system.

All the theoretical background is detailed in the paper:

> García, C.A., Otero, A., Félix, P., Presedo, J. & Márquez D.G., (2016). Simultaneous estimation of deterministic and fractal components in non-stationary time series (in review).

## The _WaveletVar_ class
According to a convenient modelling approach, long-memory signals can be 
studied as realizations of one of two classes of processes: fractional Gaussian noise (fGn) or fractional Brownian motion (fBm). Since the increments of a fBm signal yield a stationary fGn signal, we may focus on the fBm model. Thus, we
assume that $B$ can be approximated by a fBm.

Two important features should be handled when analysing fBm signals: 
non-stationarity, which demands for a time-dependent analysis; and 
self-similarity, which demands for a scale-dependent analysis. Thus, wavelet
analysis provides a proper framework for studying fBm signals. Furthermore, it 
is possible to estimate the fBm parameters by studying how the wavelet
coefficients' variance depend on the resolution level. ___fracdet___ provides 
the `WaveletVar` class for studying the wavelet coefficients' variances.

```{r waveletVar, message=FALSE, warning = FALSE}
library('fracdet')
# parameters for the example:
set.seed(3)
nlevels = 14
n = 2 ^ nlevels
H = 0.3
# Simulate a fbm and compute its wavelet transform.
b = fbmSim(n = n, H = H)
wb = wd(b, bc = "symmetric")
# Compute the wavelet coefficients variances and plot them.
vpr = WaveletVar(wb)
plot(vpr)
```

## fBm parameter estimation
The shape and slope of the wavelet variances  depends on the Hurst exponent of
the fBm. To estimate it, __fracdet__ provides the `estimatefBm` function, which
is based on the `nls` R function:

```{r parameterEstimates}
model = estimatefBmPars(vpr,
                        use_resolution_levels = c(5:13))
# 'model' is a nls object and thus,
# we can obtain the parameter estimations with:
print(coef(model))
```

## The _Fracdet_ class
What happens when our observed signal $Y$ may be approximated as 
$Y[n] = x[n] + B[n]$?

```{r example}
# simulate a simple fbm + sinusal signal.
x = 1.25 * cospi(2 * 1:n / 10)
y = fbmSim(n = n, H = H) + x
wy = wd(y, bc = "symmetric")
# Compute the wavelet coefficients variances and plot them.
vpr = WaveletVar(wy)
plot(vpr)
```

Can you see the deviations in the wavelet coefficients' variance on level 11?
This is due to the deterministic components. However, if the deterministic 
component $x$ is a band-limited signal we can still obtain good estimates
of the fBm parameters. 

```{r parameterEstimatesII}
# Ignore resolution levels 11 and 12 to obtain the 
# fBm parameters estimates.
model = estimatefBmPars(vpr,
                        use_resolution_levels = c(5:10,13))

```

The `Fracdet` class can now be used to ease the procedure of working with a signal that may fulfil the assumptions of our fractal-deterministic model. The 
parameters required for the `Fracdet` constructor are the wavelet
transform of the observed signal,  and the fitted-model object used for
estimating the fBm parameters. 


```{r Fracdet}
# Create a Fracdet object...
fd = Fracdet(wy, model)
# ... and check the fbm parameters estimates.
print(coef(fd))
# A plot of the fitted variances may also be useful.
plot(getWaveletVar(fd))
points(getFittedWaveletVar(fd),
       col = 2,
       pch = 2)
# we could also extract the nls-fit and use the predict function.
# The nls-fit is performed in semilog-space. Thus, a transformation
# of the predicted values is required.
nls_model = getFbmPars(fd)
points(resolutionLevels(vpr),
       2 ^ predict(nls_model, newdata = data.frame(x = resolutionLevels(vpr))),
       col = 3,
       pch = 3)
```

By performing the nls-fit we have not only characterized the fBm process, but we
have also gained information about the deterministic contributions in each 
wavelet resolution level. We shall use this information to obtain an estimation 
of the deterministic signal.

## Estimate the deterministic signal
Using all the available information about the observed signal, a bayesian model 
of the distribution of the deterministic and stochastic wavelet coefficients is
built. The available information consist on:

  * The estimates of the fBm parameters (which completely characterize the
  statistical properties of the fBm signal).
  * The well-known statistical properties of the fBm signals in wavelet domain.
  * The deviations from the theoretical wavelet coefficients' variance permit the
  estimation of the energy distribution across the resolution levels of the
  deterministic signal.

The wavelet transform of the deterministic signal is then estimated by calculating
the decision rule that minimizes the posterior expected value of a squared loss
function. The `estimateDetSignal` implements all these actions and returns the
estimate of the wavelet transform of the deterministic signal:

```{r signalEstimation, cache=TRUE}
# Estimate the deterministic signal (this may take a while) taking
# into account the deviations in level 11, 12. 
wx = estimateDetSignal(fd, estimate_from = 11:12)
x_est = wr(wx)
# compare the original signal and the estimation
xlim = c(300, 450)
offset = 2
indx = xlim[[1]]:xlim[[2]]
ylim = range(c(offset + y[indx],
               x[indx],
               x_est[indx]))
plot(offset + y, xlim = xlim,
     ylim = ylim,
     type = "l")
lines(x, col = 2 , lty = 2, lwd = 2)
lines(x_est, col = 3, lty = 3, lwd = 2)
legend("topright", lty = 1:3, col = 1:3,
       legend = c("Y", "x", "x estimate"), bty = "n")
```
